Kindly Bent to Free Us

### 2019-07-04 kindly/motivation.tex
motivation: hybrid arrays

The programming style demonstrated for channels is
appropriate for purely functional code, but becomes tedious when
manipulating mutable resources.
%
To make linearity easier to use for imperative programming, we rely on
the notion of borrowing \cite{DBLP:conf/popl/BoylandR05}. As an example,
\cref{sig:array} contains an API for mutable affine arrays with a use case in \cref{ex:array}.

The intention of this API is to enable arrays to contain affine
elements. One would first use \lstinline/create n v/ to create
an (unrestricted) array of size \lstinline/n/ with some value
\lstinline/v/. This value must be unrestricted as it is duplicated to
initialize all array elements. Using the \lstinline/map/ function we
can transform the unrestricted elements into linear affine)
ones. This transformation yields a linear (affine) array that can be
processed further using \lstinline/map/ (or \lstinline/set/ if
affine). Before \lstinline/free/ing the array the elements must be
affine, again.

We distinguish several kinds of operation: constructors, destructore,
observers, and mutators. 
Constructors and destructors like \lstinline/create/ and \lstinline/free/ manipulate the whole
array. As with channels in \cref{sec:session-linearity},
\lstinline/create/ yields a linear resource which is consumed by \lstinline/free/.
During the lifetime of the array resource \lstinline/a/, we can split
off \emph{immutable borrows} \lstinline/&a/ that provide a read-only
view or \emph{mutable borrows} \lstinline/&!a/ for read-write views.
Observer functions such as \lstinline/length/ and \lstinline/get/ expect an immutable borrow
argument, mutator functions such a \lstinline/set/ expect
a mutable borrow.

Generally, each borrow is tied to a region whose lifetime is properly
contained in the lifetime of the resource. 
In a region, we can split off as many immutable borrows as we like
from a resource, but we can have only one mutable borrow. In a
subsidiary region, we can take further immutable borrows from any
borrow or we can take a mutable borrow from a mutable borrow from a
enclosing region. Borrows are confined to their regions. Inside,
immutable borrows are unrestricted (\lstinline/un/) whereas mutable
borrows are affine (\lstinline/aff/).

Furthermore, any borrow prevents access to the ``raw''  resource
as long as it is in use. In \cref{ex:array}, \cref{ex:array:get},
two immutable borrows are used in the same expression. Once this operation
is done, the borrows are not used anymore, which allows us to use
an exclusive mutable borrow in \cref{ex:array:set}.
Finally, we return the whole array in \cref{ex:array:return} because
none of the borrows are in scope anymore.

\begin{figure}
  \centering
  \begin{subfigure}{0.48\linewidth}
\begin{lstlisting}
module Array : sig
  type ('a : 'k) t : 'k

@@

%%% TBC
Naturally, \lang also checks that borrows do not escape their point
of use through {\em syntactic regions}. The example
above is reproduce in \cref{ex:array:region} with
regions explicitly delimited by \lstinline/{| ... |}/.
\lang ensures that borrows never escape the smallest enclosing
region.
For instance, On \cref{ex:array:region:get}, \lstinline/&a/ does
not escape the region since the borrow is not captured, and the
region only return integers.
Regions annotations are automatically inserted by \lang using
syntactic information such as borrows and binders.
Users can also annotate regions manually to further restrict the range
of a borrow. In practice, we have found that regions
annotations are very rarely needed in common imperative usage.

@@

\begin{figure}
  \centering
  \begin{subfigure}{0.5\linewidth}
\begin{lstlisting}
let for f a  = 
  let n = Array.length &a in
  let rec loop f a i = 
    if i < n then (
      f &!a i;
      loop f &!a (i+1)
    )
  in
  loop f &!a 0
# Borrowed m ('a Array.t) tt =>
  (tt -> int -> unit) -> tt -> unit
\end{lstlisting}
    \caption{For loop function for affine arrays}
    \label{ex:array:for}
  \end{subfigure}
\end{figure}

@@

In the last few years, new persistent data-structures
have been introduced which allow transient mutations, but discourage
non-linear uses either through degraded performances
\cite{DBLP:conf/ml/ConchonF07} or
dynamic and static checks \cite{DBLP:journals/pacmpl/Puente17}.
In particular, persistent Hash-Array-Mapped-Trie has been used with similar
API in numerous non-pure functional languages (OCaml, Clojure, \dots).
Affine types allow to formalize the performance contract between the programmer
and the library, while borrows avoid the need to thread state explicitly,
as usually required by an immutable API.
%
In this section, we show how \lang allows us to define
a safe API for persistent arrays that support both immutable and mutable features,
and how it can be used to implement a simple backtracking sudoku solver.

Our sudoku solver is implemented by maintaining a grid that represent the state of the game. In order to make backtracking easier, we want to modify the grid immutably. However, while some changes correspond to backtracking points,
some do not.
In order to improve performance, we want to implement modifications
that do not correspond to backtracking points as mutations.
Thanks to affine types and borrows, we can provide an API that guarantee correct
uses of mutable and immutable modifications to the board.
This API is represented by the signature \lstinline/HybArray/ in \cref{sig:hybarray}.
It only differs with the \lstinline/Array/ module on two points.
The mutation function is now called \lstinline/set_mut/, with
a type \lstinline/&!('a t) * int * 'a -> unit/ and
the \lstinline/set/ operation
is now of type \lstinline/&('a t) * int * 'a -> 'a t/. It
takes an \emph{immutable} borrow, which allows it to read the array, and
return a fresh hybrid array.
We showcase a very simple implementation of \lstinline/HybArray/ in
\cref{ex:cow} where hybrid arrays are represented
as regular arrays with copy-on-write for immutable
modifications.
A more elaborate implementation could use persistent arrays
\cite{DBLP:conf/ml/ConchonF07}.

\begin{figure}
  \centering
  \begin{subfigure}{0.48\linewidth}
\begin{lstlisting}
module type HybArray = sig
  type ('a : un) t : aff

  val create : ('a : un) => int -> 'a -> 'a t
  val free : 'a t -> unit

@@

  let set (a, i, x) =
    let f k =
      if k = i then x else get (a,k)
    in
    create (length a) f

@@

Thanks to this API, modifications that correspond to backtracking points
use \lstinline/set/, which allows arbitrary uses of the previous version
of the array, while other modifications uses \lstinline/set_mut/, which
forbids uses of the previous version.
This can be seen in the sudoku solver implementation in \cref{ex:sudoku}.
The board is represented as a matrix of cells (\cref{line:boardtype}).
We assume the existence of a \lstinline/Matrix/ module which uses the same API
as \lstinline/HybArray/ but with two indices.
Each cell is an integer
set that represent admissible solutions so far. The sudoku solver will
iterate on the cells and try each possible solution (\cref{line:try_solution}).
When a value is picked for the current cell, we use an immutable modification
to create a backtracking point (\cref{line:immutset}) and propagate
the changes with the \lstinline/propagate/ function. The \lstinline/propagate/
function does not need to preserve the previous
version of the board, and thus uses direct mutation.
The implementation of \lstinline/propagate/ is split into three parts
for lines, columns, and square. For concision, only \lstinline/propagate_lines/ is
presented here (\cref{line:propline}).
The typing ensures that all the mutations do not compromise
backtracking, since sharing is only allowed through
the immutable set functions, which is used for backtracking points.
In particular, the \lstinline/try_solution/ function can be used in a loop
precisely because the \lstinline/set/ function only requires an
\emph{immutable} borrow of \lstinline/new_g/.
This is allowed, since immutable borrows are unrestricted.

@@

\begin{figure}
  \centering
  \begin{minipage}[t]{0.45\linewidth}
    \lstinputlisting[linerange=1-13]{code/sukodu.affe}
  \end{minipage}\hfill
  \begin{minipage}[t]{0.54\linewidth}
    \lstinputlisting[linerange=15-28,firstnumber=last]{code/sukodu.affe}
  \end{minipage}
  \caption{Chosen excerpt from the sudoku solver}
  \label{ex:sudoku}
\end{figure}

>>>>

The functional resource-passing style demonstrated for channels in
\cref{sec:session-linearity} becomes tedious when manipulating mutable
resources like buffers or connection pools.
%
To make linearity easier to use with an imperative programming style, we rely on
the notion of borrowing \cite{DBLP:conf/popl/BoylandR05}.

As a first example, we consider an API for mutable linear arrays
(\cref{sig:array}).
The intention of this API is to safely handle manual allocation and
deallocation of arrays that may contain affine elements.
One would first use \lstinline/create n v/ to create
an (unrestricted) array of size \lstinline/n/ with some value
\lstinline/v/. This value must be unrestricted as it is duplicated to
initialize all array elements. Using the \lstinline/map/ function we
can transform the unrestricted elements into linear (affine)
ones. This array can be processed further using \lstinline/map/ (or \lstinline/set/ if
affine). The \lstinline/get/ function only applies if the element type
is unrestricted as one element is duplicated. The \lstinline/length/
function applies anytime. Before \lstinline/free/ing the array the elements must be
affine, again.

To manage the different kinds of accessing the array we distinguish between constructors, destructors, 
observers, and mutators. 
Constructors and destructors like \lstinline/create/ and \lstinline/free/ manipulate the whole
array. As with channels in \cref{sec:session-linearity},
\lstinline/create/ yields a linear resource which is consumed by \lstinline/free/.
During the lifetime of the array resource \lstinline/a/, we can split
off \emph{immutable borrows} \lstinline/&a/ that provide a read-only
view or \emph{mutable borrows} \lstinline/&!a/ for read-write views.
Observer functions such as \lstinline/length/ and \lstinline/get/ expect an immutable borrow
argument, mutator functions such a \lstinline/set/ expect
a mutable borrow.

Each borrow is tied to a region whose lifetime is properly
contained in the lifetime of the resource. 
In a region, we can split off as many immutable borrows as we like
from a resource, but we can take only one mutable borrow. In a
subsidiary region, we can take further immutable borrows from any
borrow or we can take a mutable borrow from a mutable borrow from a
enclosing region. Borrows are confined to their regions. Inside the region,
immutable borrows are unrestricted (\lstinline/un/) whereas mutable
borrows are affine (\lstinline/aff/).

The code in \cref{ex:array} demonstrates a use of the API to create an
array of Fibonacci numbers. After creation of the array, we can see
that the presence of a borrow prevents access to the ``raw'' 
resource inside the borrow's region. In \cref{ex:array:get} of the
example, there are two immutable borrows in the same
expression, which forms a region by itself. These borrows are split
off the mutable borrow used in \cref{ex:array:set}, which belongs to
the next enclosing region.
Finally, the code returns the whole array in \cref{ex:array:return},
which is possible because none of the borrows are in scope anymore.

\begin{figure}[tp]
  \centering
  \begin{subfigure}{0.48\linewidth}
\begin{lstlisting}
module Array : sig
  type ('a : 'k) t : lin

@@

\cref{ex:array:region} makes explicit the regions from the example
indicated by braces as in \lstinline/{| ... |}/.  \lang guarantees that borrows
never escape the smallest enclosing region.  For instance, in
\cref{ex:array:region:get}, \lstinline/&a/ does not escape because it
is consumed in the calls to \lstinline/get/.
Regions annotations are automatically inserted using
syntactic information such as borrows and binders (see
\cref{regionannot} for more information).  Users can also
annotate regions manually to further restrict the range of a
borrow. In practice, we have found that region annotations are
rarely needed with common programming idioms. All regions for the
examples in this section have been inferred.

@@

\begin{figure}
  \centering
  \begin{subfigure}{0.5\linewidth}
\begin{lstlisting}
let for f a  = 
  let n = Array.length &a in
  let rec loop f a i = 
    if i < n then (
      f &!a i;
      loop f &!a (i+1)
    )
  in
  loop f &!a 0
# for : Borrowed m ('a Array.t) tt =>
  (tt -> int -> unit) -> tt -> unit
\end{lstlisting}
    \caption{For loop function for affine arrays}
    \label{ex:array:for}
  \end{subfigure}
\end{figure}

@@

In the last few years, new persistent data structures
have been introduced which permit transient mutations, but discourage
non-linear uses either through degraded performance
\cite{DBLP:conf/ml/ConchonF07} or
dynamic and static checks \cite{DBLP:journals/pacmpl/Puente17}.
In particular, persistent Hash-Array-Mapped-Tries have been used with similar
APIs in several non-pure functional languages (OCaml, Clojure, \dots).
Affine types allow us to formalize the performance contract between the programmer
and the library, while borrows avoid the need to thread state explicitly,
as usually required by an API that enforces immutability.
%
In this section, we show how \lang allows us to define
a safe API for persistent arrays that support both immutable and mutable features,
and how it can be used to implement a simple backtracking Sudoku solver.

The Sudoku solver maintains a 2D-array that represents the state of
the game and uses backtracking when there are several alternatives to proceed.
The use of backtracking suggests a persistent data structure for the 2D-array.
However, only changes that correspond to a choice point need to use
the persistence mechanism, others may be implemented as
cheap in-place mutations.

\cref{sig:hybarray} contains an API \lstinline/HybArray/ 
that guarantees correct use of mutable and immutable modifications to
the board by relying on affine types and borrows.
The signature differs slightly from the \lstinline/Array/
signature. For simplicity, the array elements are unrestricted, 
hence it is sufficent for the array to be affine.
The in-place mutation function is called \lstinline/set_mut/ with type
\lstinline/&!('a t) * int * 'a -> unit/ and the persistent
\lstinline/set/ operation has type \lstinline/&('a t) * int * 'a -> 'a t/. It
takes an \emph{immutable borrow} because it only reads from the
argument array, but returns a fresh,  modified structure.
The code in \cref{ex:cow} is a very simple implementation of
\lstinline/HybArray/ that represents hybrid arrays
as regular arrays with copy-on-write for persistent
modifications. The function
\lstinline/create_from : int -> (int -> 'a) -> 'a Array.t /
initializes an array from a function.
% A more elaborate implementation could use persistent arrays
% \cite{DBLP:conf/ml/ConchonF07}.

\begin{figure}[tp]
  \centering
  \begin{subfigure}{0.48\linewidth}
\begin{lstlisting}
module type HybArray = sig
  type ('a : un) t : aff

  val create : int -> 'a -> 'a t
  val free : 'a t -> unit

@@

  let set (a, i, x) =
    let f k =
      if k = i then x else get (a,k)
    in
    create_from (length a) f

@@

Thanks to this API, modifications that correspond to backtracking points
use \lstinline/set/, which allows arbitrary uses of the previous version
of the array, while other modifications uses \lstinline/set_mut/, which
forbids uses of the previous version.
This can be seen in the Sudoku solver implementation in \cref{ex:sudoku}.
The board is represented as a matrix of cells (\cref{line:boardtype}).
We assume the existence of a \lstinline/Matrix/ module which uses the same API
as \lstinline/HybArray/ but with two indices.
Each cell is an integer
set that represent admissible solutions so far. The Sudoku solver will
iterate on the cells and try each possible solution (\cref{line:try_solution}).
When a value is picked for the current cell, we use an immutable modification
to create a backtracking point (\cref{line:immutset}) and propagate
the changes with the \lstinline/propagate/ function. The \lstinline/propagate/
function does not need to preserve the previous
version of the board, and thus uses direct mutation.
The implementation of \lstinline/propagate/ is split into three parts
for lines, columns, and square. For concision, only \lstinline/propagate_lines/ is
presented here (\cref{line:propline}).
The typing ensures that all the mutations do not compromise
backtracking, since sharing is only allowed through
the immutable set functions, which is used for backtracking points.
In particular, the \lstinline/try_solution/ function can be used in a loop
precisely because the \lstinline/set/ function only requires an
\emph{immutable} borrow of \lstinline/new_g/.
This is allowed, since immutable borrows are unrestricted.

@@

\begin{figure}
  \centering
  \begin{minipage}[t]{0.45\linewidth}
    \lstinputlisting[linerange=1-13]{code/sukodu.affe}
  \end{minipage}\hfill
  \begin{minipage}[t]{0.54\linewidth}
    \lstinputlisting[linerange=15-28,firstnumber=last]{code/sukodu.affe}
  \end{minipage}
  \caption{Excerpt of the Sudoku solver}
  \label{ex:sudoku}
\end{figure}
